---
title: 'StenoScope: Author Graph'
author: "Tommi Suvitaival, Steno Diabetes Center Copenhagen, tommi.raimo.leo.suvitaival@regionh.dk"
date: "`r Sys.Date()`"
output: 
  html_document:
    keep_md: true
    fig_width: 9
    fig_height: 9
    dev: jpeg
    toc: yes
  github_document:
    fig_width: 9
    fig_height: 9
    dev: jpeg
    toc: yes
always_allow_html: true
---

# Introduction

This is a tutorial script to producing an **interactive visualization of a co-author network**. The network is based on a publication list of Steno Diabetes Center Copenhagen in 2021, extracted from [the Capital Region of Denmark](https://research.regionh.dk/en/publications/search.html?search=&organisationName=Steno%20Diabetes%20Center%20Copenhagen&publicationYearsFrom=2021&journalName=&organisations=49636648&publicationstatus=published&publicationcategory=33079909&peerreview=true&language=%20&publicationYearsTo=2021&type=%2Fdk%2Fatira%2Fpure%2Fresearchoutput%2Fresearchoutputtypes%2Fcontributiontojournal%2F%25&uri=&pageSize=500&page=0). Each author is a node in the network, and the author's position and vertices (connections to other authors) are dependent on, whom the author has co-authored publications with. Moreover, the total number of publications and co-author of the author are shown as additional information, when hovering the cursor over the author.

The code is for R and it is, among others, based on packages [stringr](https://stringr.tidyverse.org/), [ggplot2](https://ggplot2.tidyverse.org/) and [plotly](https://plotly.com/r/).

This document is part of the Github repository [StenoScope](https://github.com/tommi-s/StenoScope) by [Tommi Suvitaival](https://tommi-s.com/).

View this document at https://tommi-s.com/StenoScope/AuthorGraph/ to show all output correctly.

```{r, echo = FALSE}

# Load the previously-generated result to show the final result already in the beginning.

path.result <- paste0( here::here(), "/output/AuthorGraph/AuthorGraph.rds" )

if ( file.exists( path.result ) ) {

  plot.interactive.loaded <- readRDS( path.result )
  
  plot.interactive.loaded
  
}

```

# Load Data

* The data set has been created by copying the contents from a publication list website to a text file.
* Source: [Publications from Steno Diabetes Center Copenhagen in 2021](https://research.regionh.dk/en/publications/search.html?search=&organisationName=Steno%20Diabetes%20Center%20Copenhagen&publicationYearsFrom=2021&journalName=&organisations=49636648&publicationstatus=published&publicationcategory=33079909&peerreview=true&language=%20&publicationYearsTo=2021&type=%2Fdk%2Fatira%2Fpure%2Fresearchoutput%2Fresearchoutputtypes%2Fcontributiontojournal%2F%25&uri=&pageSize=500&page=0) (the Capital Region of Denmark; accessed 1.2.2022).

```{r Load-Data}

file.path <-
  paste0(
    here::here(),
    "/data/publications-SDCC-2021.txt"
  )

data.loaded <- 
  read.delim(
    file = file.path, # "../data/publications-SDCC-2021.txt"
    encoding = "UTF-8",
    header = FALSE,
    stringsAsFactors = FALSE
  )

```

# Prepare the Data

## Create a Data Frame

* Create a data frame with publications as rows and properties as columns.

```{r Data-Frame}

data <-
  matrix(
    data = unlist( data.loaded ),
    ncol = 4,
    byrow = TRUE
  )

colnames( data ) <-
  c(
    "Status",
    "Title",
    "Info",
    "Type"
  )

data <- 
  data.frame(
    data,
    stringsAsFactors = FALSE
  )

head( data )

```

## Subset by Publication Type

* Include journal articles and review articles in the analysis.

```{r Subset-Publications}

table( data$"Type" )

is.included <- 
  grepl( 
    x = data$"Type",
    pattern = "(Journal article)|(Review)"
  )

data <- data[ is.included, ]

```

## Create Additional Properties

* Separate authors from the information column, which also contains the publication date and the journal reference.

```{r Properties}

# Split by the date text.

tmp <-
  stringr::str_split_fixed(
    string = data[ , 3 ],
    pattern = "(\\,\\s([0-9]+\\s)?([A-Z][a-z][a-z]\\s)?2021\\,\\s)",
    n = 2
  )

colnames( tmp ) <- 
  c(
    "Authors",
    "Reference"
  )

tmp <- 
  data.frame(
    tmp,
    stringsAsFactors = FALSE
  )

data <-
  dplyr::bind_cols(
    data,
    tmp
  )

str( data )

```

## Split Individual Authors

* Separate individual authors, who are joined by comma or &-sign.
* This creates a list, where each item corresponds to a publication, and the contents of the item is a character vector containing the names of the authors in the respective publication.

```{r Authors-Split}

authors <-
  stringr::str_split(
    string = data$"Authors",
    pattern = "(\\.\\, )|(\\. \\& )|( \\& )"
  )

head( authors )

```

## Format Author Names

* Remove additional punctuations and whitespaces.

```{r Authors-Format}

authors <-
  lapply(
    X = authors,
    FUN = stringr::str_remove_all,
    pattern = "\\,"
  )

authors <-
  lapply(
    X = authors,
    FUN = stringr::str_remove_all,
    pattern = "\\. "
  )

authors <-
  lapply(
    X = authors,
    FUN = stringr::str_remove_all,
    pattern = "\\."
  )

authors <-
  lapply(
    X = authors,
    FUN = stringr::str_trim,
    side = "both"
  )

authors <-
  lapply(
    X = authors,
    FUN = stringr::str_squish
  )

head( authors )

```

## Extract unique authors

```{r Authors-Unique}

authors.unique <- sort( x = unique( unlist( authors ) ) )

head( authors.unique )

```

# Network of Authors

## Initialize the Adjacency Matrix

* Authors-by-authors Adjacency matrix

```{r Adjacency-Initialize}

adjacency.mat <-
  array( 
    data = 0,
    dim = c( 1, 1 ) * length( authors.unique )
  )

rownames( adjacency.mat ) <-
  colnames( adjacency.mat ) <-
  authors.unique

str( adjacency.mat )

str( 
  object = adjacency.mat, 
  vec.len = 5
)

```

## Compute the Co-Occurrence of Author Pairs

* Go through all publications 'i'.
* Add up the co-occurrence count for the authors in the publication 'i'.

```{r Adjacency-Compute}

for ( i in 1:length( authors ) ) {
  
  tmp <- rownames( adjacency.mat ) %in% authors[[ i ]]
  
  adjacency.mat[ tmp, tmp ] <- adjacency.mat[ tmp, tmp ] + 1
  
}

str( adjacency.mat )

```

## Subset Authors

* Include authors, who have more than one publication.

```{r Subset-Authors}

is.included <- diag( adjacency.mat ) > 1

data.plot <- adjacency.mat[ is.included, is.included ]

str( data.plot )

```

## Wrap Author Names

* Add line break upon whitespace.
* Add line break upon dash.

```{r Authors-Wrap}

names <- rownames( data.plot )

names <-
  stringr::str_replace_all(
    string = names,
    pattern = "\\s",
    replacement = "\n"
  )

names <-
  stringr::str_replace_all(
    string = names,
    pattern = "\\-",
    replacement = "-\n"
  )

names <-
  stringr::str_replace_all(
    string = names,
    pattern = "\\n\\-",
    replacement = ""
  )

names( names ) <- rownames( data.plot )

head( names )

```

## Create a Network Representation of the Adjacency Matrix

* Create a network representation of the adjacency matrix with the **network** package.
* The network should be non-directional.

```{r Network}

net.network <- 
  network::network(
    x = data.plot,
    directed = FALSE
  )

```

## Lay Out the Network

* Lay out the network on a two-dimensional plane (i.e., page) using the **ggnetwork** package.
* The created data frame contains the definition of the network layout, where each row is a vertex:
  + *x* and *y* are the starting coordinates of the vertex
  + *xend* and *yend* are the ending coordinates of the vertex
  + *vertex.names* is the name of the node in the starting coordinates
  + (in an undirected graph, starting and ending nodes do not have topological difference)

```{r Layout}

ggnetwork <- ggnetwork::ggnetwork( x = net.network )

head( ggnetwork )

```

## Define Additional Properties for the Visualization

* Define properties of the network:
  + Degree: number of vertices (i.e., connections) from each node (here: number of publications for each author)
  + Size: size of each node in the visualization (here: based on degree)
  + Label: Name of each node (here: author) to show in the visualization
  + Text: Additional information to show for each node on hover
    + Name of the node (here: author)
    + Degree of the node (here: number of publications)
    + Number of neighbors to the node (here: number of co-authors)

```{r Visualization-Properties}

ggnetwork$"degree" <- diag( data.plot )[ ggnetwork$"vertex.names" ]

ggnetwork$"size" <- ggnetwork$"degree" * 10

ggnetwork$"name.formatted" <- names[ ggnetwork$"vertex.names" ]

# Compute the number of co-authors, 
# including co-authors, who might be outside the subset.

tmp <- rowSums( adjacency.mat > 0 ) - 1
# tmp <- rowSums( data.plot > 0 ) - 1

ggnetwork$"N.neighbors" <- tmp[ ggnetwork$"vertex.names" ]

ggnetwork$"text" <-
  paste0(
    ggnetwork$"vertex.names",
    ":\n\t",
    ggnetwork$"degree",
    " publications",
    "\n\t",
    ggnetwork$"N.neighbors",
    " co-authors"
  )

head( ggnetwork )

```

## Create the Basic Visualization

* Using the **ggplot2** package with rendering of the nodes and edges from the **ggnetwork** package.

```{r Figure-Network-Basic}

plot <-
  ggplot2::ggplot(
    data = ggnetwork,
    mapping =
      ggplot2::aes(
        x = x,
        y = y,
        xend = xend,
        yend = yend
      )
  ) +
  ggnetwork::geom_edges() +
  ggnetwork::geom_nodes()

plot

```

## Create (the Passive) Visualization with Additional Properties

* Add additional properties:
  + Text for tooltip (used later in the interactive version)
  + Formatted names of the nodes (here: authors)
  + Transparent vertices
  + Node size according to the degree of the node (here: number of publications)
  + Simple white colour theme

```{r Figure-Network}

plot <-
  ggplot2::ggplot(
    data = ggnetwork,
    mapping =
      ggplot2::aes(
        x = x,
        y = y,
        xend = xend,
        yend = yend,
        label = name.formatted,
        text = text
      )
  ) +
  ggnetwork::geom_edges( alpha = 0.125 ) +
  ggnetwork::geom_nodes(
    mapping =
      ggplot2::aes(
        size = size
      )
  ) +
  ggnetwork::geom_nodetext( 
    alpha = 0.25,
    size = 3
  ) +
  ggthemes::theme_solid()

plot

```

## Create the Interactive Visualization

* Create an interactive version of the visualization with the **ggplotly** function from the **plotly** package.
* Show info text about an author as tooltip on hover-over.

```{r Figure-Network-Interactive-Create}

plot.interactive <-
  plotly::ggplotly(
    p = plot,
    tooltip = c( "text" )
  )

```

## View the Interactive Visualization

```{r Figure-Network-Interactive}

plot.interactive

```

## Save the Result

* (Uncomment for saving the files.)

```{r Figure-Network-Interactive-Export}

# Save result to "../data/publications-SDCC-2021.txt"
path.result <- paste0( here::here(), "/output/AuthorGraph/" )

# dir.create(
#   path = path.result,
#   recursive = TRUE
# )
# 
# saveRDS(
#   object = plot.interactive,
#   file = paste0( path.result, "AuthorGraph.rds" )
# )

```

# SessionInfo

```{r SessionInfo}

utils::sessionInfo()

```

# Appendix

* Copy README.html to index.html to show this document on Github Pages.

```{r index-html}

if ( file.exists( "README.html" ) ) {
  
  system( command = "rm index.html" )
  system( command = "cp README.html index.html" )
  
}

```